---
title: "Dayton Weather Graph"
output: html_notebook
---

Dayton 2014 Weather Graph - replication
https://rpubs.com/bradleyboehmke/weather_graphic

```{r}
# clear environment
rm(list=ls())
```

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(data.table)
```

https://www.r-bloggers.com/getting-data-from-an-online-source/

```{r}
DAY <- fread('http://academic.udayton.edu/kissock/http/Weather/gsod95-current/OHDAYTON.txt')
```

```{r}
names(DAY) <- c("Month", "Day", "Year", "Temp")
```

```{r}
# Historical Temperature - calculate historical min, max, and normal range
Past <- DAY %>%
  #group_by(Year, Month) %>%
  #arrange(Day) %>%
  #ungroup() %>%
  group_by(Year) %>%
  mutate(dayOfYear = seq(1, length(Day))) %>% # x-axis labels
  ungroup() %>%
  
  # filter missing data and years 2014 - 2018
  filter(Temp != -99 & Year != 2014 & Year != 2015 & Year != 2016 & Year != 2017 & Year != 2018) %>% 
  group_by(dayOfYear) %>%
  mutate(upper = max(Temp),
         lower = min(Temp),
         avg = mean(Temp),
         se = sd(Temp)/sqrt(length(Temp))) %>%
  mutate(avg_upper = avg + (2.101 * se), 
         avg_lower = avg - (2.101 * se))  %>%
  ungroup()
```

```{r}
head(Past)
tail(Past)
glimpse(Past)
```

```{r}
Present <- DAY %>%
  #group_by(Year, Month) %>%
  #arrange(Day) %>%
  #ungroup() %>%
  group_by(Year) %>%
  mutate(dayOfYear = seq(1, length(Day))) %>%
  ungroup() %>%
  filter(Temp != -99 & Year == 2014)
```

```{r}
# create dataframe that represents the lowest temp for each day for the historical data
PastLows <- Past %>%
        group_by(dayOfYear) %>%
        summarise(Pastlow = min(Temp)) # identify lowest temp for each day from 1995-2013

# create dataframe that identifies the days in 2014 in which the temps were lower than all previous 19 years
PresentLows <- Present %>%
        left_join(PastLows) %>%  # merge historical lows to current year low data
        mutate(record = ifelse(Temp<Pastlow, "Y", "N")) %>% # identifies if current year was record low
        filter(record == "Y")  # filter for days that represent current year record lows

# create dataframe that represents the highest temp for each day for the historical data
PastHighs <- Past %>%
        group_by(dayOfYear) %>%
        summarise(Pasthigh = max(Temp))  # identify highest temp for each day from 1995-2013

# create dataframe that identifies the days in 2014 in which the temps were higher than all previous 19 years
PresentHighs <- Present %>%
        left_join(PastHighs) %>%  # merge historical highs to current year low data
        mutate(record = ifelse(Temp>Pasthigh, "Y", "N")) %>% # identifies if current year was record high
        filter(record == "Y")  # filter for days that represent current year record highs
```

```{r}
# function to turn y-axis labels into degree formatted values
dgr_fmt <- function(x, ...) {
        parse(text = paste(x, "*degree", sep = ""))
}

# create y-axis variable
a <- dgr_fmt(seq(-20, 110, by=10))

# create a small dataframe to represent legend symbol for 2014 Temperature
legend_data <- data.frame(x=seq(175, 182),y=rnorm(8, 15, 2))
```

geom_linerange() vertical intervals: lines, crossbars, and errorbars defined by x, ymin, xmax

geom_linerange(mapping = NULL, data = NULL, stat = "identity",
  position = "identity", ..., na.rm = FALSE, show.legend = NA,
  inherit.aes = TRUE)

```{r}
# Historical Temp high/low
p <- ggplot(Past, aes(dayOfYear, Temp)) + 
  theme(plot.background = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.ticks = element_blank(),
        # axis.text = element_blank(),
        axis.title = element_blank()) +
  geom_linerange(aes(x=dayOfYear, ymin=lower, ymax=upper), colour = "wheat2", alpha = 0.1) +
  geom_vline(xintercept = 0, colour = "wheat4", linetype = 1, size = 1)
print(p)
```

```{r}
# 95% CI
p <- p +
  geom_linerange(Past, mapping = aes(x=dayOfYear, ymin=avg_lower, ymax=avg_upper), colour = "wheat4")
print(p)
```

```{r}
# 
p <- p + 
  geom_line(Present, mapping=aes(x=dayOfYear, y=Temp, group=1))
print(p)
```

```{r}
# y-axis white gridlines
p <- p +
        geom_hline(yintercept = -20, colour = "white", linetype=1) +
        geom_hline(yintercept = -10, colour = "white", linetype=1) +
        geom_hline(yintercept = 0, colour = "white", linetype=1) +
        geom_hline(yintercept = 10, colour = "white", linetype=1) +
        geom_hline(yintercept = 20, colour = "white", linetype=1) +
        geom_hline(yintercept = 30, colour = "white", linetype=1) +
        geom_hline(yintercept = 40, colour = "white", linetype=1) +
        geom_hline(yintercept = 50, colour = "white", linetype=1) +
        geom_hline(yintercept = 60, colour = "white", linetype=1) +
        geom_hline(yintercept = 70, colour = "white", linetype=1) +
        geom_hline(yintercept = 80, colour = "white", linetype=1) +
        geom_hline(yintercept = 90, colour = "white", linetype=1) +
        geom_hline(yintercept = 100, colour = "white", linetype=1) +
        geom_hline(yintercept = 110, colour = "white", linetype=1)

print(p)
```

```{r}
# x-axis dotted gridlines
p <- p + 
        geom_vline(xintercept = 31, colour = "wheat4", linetype=3, size=.5) +
        geom_vline(xintercept = 59, colour = "wheat4", linetype=3, size=.5) +
        geom_vline(xintercept = 90, colour = "wheat4", linetype=3, size=.5) +
        geom_vline(xintercept = 120, colour = "wheat4", linetype=3, size=.5) +
        geom_vline(xintercept = 151, colour = "wheat4", linetype=3, size=.5) +
        geom_vline(xintercept = 181, colour = "wheat4", linetype=3, size=.5) +
        geom_vline(xintercept = 212, colour = "wheat4", linetype=3, size=.5) +
        geom_vline(xintercept = 243, colour = "wheat4", linetype=3, size=.5) +
        geom_vline(xintercept = 273, colour = "wheat4", linetype=3, size=.5) +
        geom_vline(xintercept = 304, colour = "wheat4", linetype=3, size=.5) +
        geom_vline(xintercept = 334, colour = "wheat4", linetype=3, size=.5) +
        geom_vline(xintercept = 365, colour = "wheat4", linetype=3, size=.5) 

print(p)
```

```{r}
# axis labels
p <- p +
        coord_cartesian(ylim = c(-20,110)) +
        scale_y_continuous(breaks = seq(-20,110, by=10), labels = a) +
        scale_x_continuous(expand = c(0, 0), 
                           breaks = c(15,45,75,105,135,165,195,228,258,288,320,350),
                           labels = c("Jan", "Feb", "Mar", "Apr",
                                      "May", "Jun", "Jul", "Aug", "Sep",
                                      "Oct", "Nov", "Dec"))

print(p)
```

```{r}
p <- p + 
  geom_point(data=PresentLows, aes(x=dayOfYear, y=Temp), colour="blue3") +
  geom_point(data=PresentHighs, aes(x=dayOfYear, y=Temp), colour="firebrick3")

print(p)
```

```{r}
# title
p <- p +
    labs(title = "Dayton's Weather in 2014", subtitle = "Temperature")

print(p)
```

## Create an annotation layer
properties of the geoms are passed in as vectors.

annotate(geom, x = NULL, y = NULL, xmin = NULL, xmax = NULL,
  ymin = NULL, ymax = NULL, xend = NULL, yend = NULL, ...,
  na.rm = FALSE)

```{r}
# Annotation Legend
p <- p +
  theme(plot.title=element_text(face="bold",hjust=.012,vjust=.8,colour="#3C3C3C",size=20)) +
        annotate("text", x = 19, y = 98, label = "", size=4, fontface="bold") +

        annotate("text", x = 15, y = 113, label = "Legend", 
                 size=3, colour="gray30") +
        annotate("text", x = 62, y = 89, 
                 label = "", 
                 size=3, colour="gray30") +
        annotate("text", x = 64, y = 85, 
                 label = "", 
                 size=3, colour="gray30") +
        annotate("text", x = 18, y = 81, label = "", 
                 size=3, colour="gray30")

print(p)
```

```{r}
# more annotations
p <- p +
        annotate("segment", x = 30, xend = 40, y = -5, yend = -10, colour = "blue3") +
        annotate("text", x = 65, y = -10, label = "We had 35 days that were the", size=3, colour="blue3") +
        annotate("text", x = 56, y = -14, label = "coldest since 1995", size=3, colour="blue3") +
        annotate("segment", x = 302, xend = 307, y = 74, yend = 82, colour = "firebrick3") +
        annotate("text", x = 333, y = 82, label = "We had 19 days that were the", size=3, colour="firebrick3") +
        annotate("text", x = 324, y = 78, label = "hottest since 1995", size=3, colour="firebrick3")

print(p)
```


```{r}
# Legend
p <- p +
        annotate("segment", x = 181, xend = 181, y = 5, yend = 25, colour = "wheat2", size=3) +
        annotate("segment", x = 181, xend = 181, y = 12, yend = 18, colour = "wheat4", size=3) +
        geom_line(data=legend_data, aes(x=x,y=y)) +
        annotate("segment", x = 183, xend = 185, y = 17.7, yend = 17.7, colour = "wheat4", size=.5) +
        annotate("segment", x = 183, xend = 185, y = 12.2, yend = 12.2, colour = "wheat4", size=.5) +
        annotate("segment", x = 185, xend = 185, y = 12.2, yend = 17.7, colour = "wheat4", size=.5) +
        annotate("text", x = 196, y = 14.75, label = "NORMAL RANGE", size=2, colour="gray30") +
        annotate("text", x = 162, y = 14.75, label = "2014 TEMPERATURE", size=2, colour="gray30") +
        annotate("text", x = 193, y = 25, label = "RECORD HIGH", size=2, colour="gray30") +
        annotate("text", x = 193, y = 5, label = "RECORD LOW", size=2, colour="gray30")

print(p)

```


